MCC Models

### 2019-08-05 papers/full_paper/SI_appendix.Rmd
finish speaker probabilities calculations

We assume that the intentional goal of the speaker is to get the listener to get an option of the correct type. That is, the informativity of an utterance is calculated with respect to conveying the correct object type.
First, we calculate the literal listener's posterior distribution over referents and lexica, and marginalize (average) over objects of the same type and the different lexica. 
The literal listener's posterior distribution over referents and lexica has support of size six, owing to the unique combinations of referents and lexica.
$$
\begin{aligned}
& P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^1, u_3) P(r_1^r) P(L_1) = 0 \times 0.26 \times 0.5 =0\\
& P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^2, u_3) P(r_2^r) P(L_1) = 1 \times 0.26 \times 0.5 =0.13\\
& P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_y^2, u_3) P(r_3^y) P(L_1) = 0 \times 0.48 \times 0.5 =0\\
& P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^1, u_3) P(r_1^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^2, u_3) P(r_2^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_y^2, u_3) P(r_3^y) P(L_2) = 1 \times 0.48 \times 0.5=0.24\\
\end{aligned}
$$

@@

$$
\begin{aligned}
& P_{L_0}(r_p, \mathcal{L}_1 | {u_3}) = P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  + P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  = 0+ 0.35 =0.35 \\
& P_{L_0}(r_y, \mathcal{L}_1 | {u_3}) = P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  = 0 \\
& P_{L_0}(r_p, \mathcal{L}_2 | {u_3})  = P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  + P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  = 0+ 0 =0 \\
& P_{L_0}(r_y, \mathcal{L}_2 | {u_3}) = P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3}) = 0.65\\
\end{aligned}
$$

@@

$$
\begin{aligned}
&P_{L_0}(r_p|u_3) =  P_{L_0}(r_p, \mathcal{L}_1|u_3) + P_{L_0}(r_p, \mathcal{L}_2|u_3) = 0.35+0=0.35 \\
&P_{L_0}(r_y|u_3) =  P_{L_0}(r_y, \mathcal{L}_1|u_3) + P_{L_0}(r_y, \mathcal{L}_2|u_3) = 0+0.65=0.65 \\
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_3; r_p) = log(0.35) = -1.04 \\
&Informativity(u_3; r_y) = log(0.65) = -0.43
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_1; r_p) = log(P_{L_0}(r_p|u_1)) = log(1)  = 0\\
&Informativity(u_1; r_y) = log(P_{L_0}(r_y|u_1)) = log(0) = -\infty
\end{aligned}
$$
We assume that the speaker knows the lexicon, but doesn't believe the listener to know the lexicon. That is, the generative process of the utterance is tantamount to pointing to the table with the referent of the type they want and incidentally labeling it. The label itself carries no information.

We can now compute the probability with each a speaker would produce each of the 4 utterances given the goal of conveying each of the 2 types.

$$
P_{S_1}(u|r) = \frac{\exp(\alpha \cdot Informativity(u;r))}{\sum_{u'} \exp(\alpha \cdot Informativity(u';r))}
$$

We begin with $r_p$. If the lexicon was $\mathcal{L}_1$, the speaker would say ...

$$
P_{S_1}(u|r^1_p) = \frac{\exp(\alpha \cdot Informativity(u;r_p))}{\sum_{u'} \exp(\alpha \cdot Informativity(u';r_p))}
$$
$$
\exp(\alpha \cdot Informativity(u;r_p)) = \begin{cases} 
\exp(2.24 \times 0) = 1 & \mbox{for } u_1 \\
\exp(2.24 \times 0) = 1& \mbox{for } u_2 \\
\exp(2.24 \times -\infty) = 0  & \mbox{for } u_3 \\
\exp(2.24 \times -\infty) = 0& \mbox{for } u_4 \\
\end{cases}
$$

To arrive at the speaker production probabilities, we divide by the normalizing constant (the sum of all of the unnormalized probabilties):

$$
P_{S_1}(u|r^1_p) = \begin{cases} 
0.5  & \mbox{for } u_1 \\
0.5 & \mbox{for } u_2 \\
 0  & \mbox{for } u_3 \\
0& \mbox{for } u_4 \\
\end{cases}
$$

For $r^2_p$, we have:

$$
\exp(\alpha \cdot Informativity(u;r^2_p)) = \begin{cases} 
\exp(2.24 \times -\infty) = 0 & \mbox{for } u_1 \\
\exp(2.24 \times -\infty) = 0& \mbox{for } u_2 \\
\exp(2.24 \times -1.04) = 0.1  & \mbox{for } u_3 \\
\exp(2.24 \times -1.04) = 0.1 & \mbox{for } u_4 \\
\end{cases}
$$

and so:

$$
P_{S_1}(u|r^1_p) = \begin{cases} 
0 & \mbox{for } u_1 \\
0 & \mbox{for } u_2 \\
0.5  & \mbox{for } u_3 \\
0.5 & \mbox{for } u_4 \\
\end{cases}
$$

Next we can insert these values into equation 2 to compute the probability that the speaker chooses this utterance to refer to this type of object. To do so, we need the speaker optimality parameter of $\alpha = 2.24$, which was estimated based on Experiment 1.

TO CREATE S(u | r), WE NEED TO HAVE THE INFORMATIVITY OF EACH OF THE 4 UTTERANCES. PASSING THE INFORMATIVITY THROUGH THE SOFT-MAX (speaker optimality) creates an "unnormalized" probability, which we then turn into a real probability by dividing by the sum of all the unnormalized probabilities of each utterance.

$$
\exp(\alpha Informativity(u_3;r)) = \exp(2.24 \times)
$$

SEEMS LIKE WE NEED TO NORMALIZE ACROSS ALL UTTERANCES, SO WE WILL NEED TO REDO THE ABOVE CALCULATIONS FOR EACH OF THE 6 UTTERANCES.

$$
\begin{aligned}
&P_{S_1}(u_3|r_2) = \frac{exp(2.24 \cdot -1.03)}{exp(2.24 \cdot -1.03) + exp(2.24 \cdot 0)} = 0.09 \\
&P_{S_1}(u_3|r_3) = \frac{exp(2.24 \cdot -0.44)}{exp(2.24 \cdot -0.44)} = 1 \\
\end{aligned}
$$

Here, the informativity for a referent is normalized across all utterances that could be used to refer to this referent.

Finally, based on equation 1 we can use these values to compute the probability that the speaker thinks that the speaker is referring to the yellow object ($r_1$) when they produce $r_1$:

$$
\begin{aligned}
&P_{L_1}(r_1,L|u_1) = \frac{1 * 0.64}{1\cdot0.64 + 0.09\cdot0.36 + 0.09\cdot0.36} = 0.91
\end{aligned}
$$

>>>>

We assume that the intentional goal of the speaker is to get the listener to get an option of the correct type. That is, the informativity of an utterance is calculated with respect to conveying the correct object type as opposed to a particular referent (token).
First, we calculate the literal listener's posterior distribution over referents and lexica, and marginalize (average) over objects of the same type and the different lexica to compute the informativity of an utterance for a type.
The literal listener's posterior distribution over referents and lexica has support of size six, owing to the unique combinations of referents and lexica.
$$
\begin{aligned}
& P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^1, u_3) P(r_1^r) P(L_1) = 0 \times 0.26 \times 0.5 =0\\
& P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^2, u_3) P(r_2^r) P(L_1) = 1 \times 0.26 \times 0.5 =0.13\\
& P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_y^2, u_3) P(r_3^y) P(L_1) = 0 \times 0.48 \times 0.5 =0\\
& P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^1, u_3) P(r_1^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^2, u_3) P(r_2^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_y^2, u_3) P(r_3^y) P(L_2) = 1 \times 0.48 \times 0.5=0.24\\
\end{aligned}
$$

@@

$$
\begin{aligned}
& P_{L_0}(t_p, \mathcal{L}_1 | {u_3}) = P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  + P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  = 0+ 0.35 =0.35 \\
& P_{L_0}(t_y, \mathcal{L}_1 | {u_3}) = P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  = 0 \\
& P_{L_0}(t_p, \mathcal{L}_2 | {u_3})  = P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  + P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  = 0+ 0 =0 \\
& P_{L_0}(t_y, \mathcal{L}_2 | {u_3}) = P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3}) = 0.65\\
\end{aligned}
$$

@@

$$
\begin{aligned}
&P_{L_0}(t_p|u_3) =  P_{L_0}(t_p, \mathcal{L}_1|u_3) + P_{L_0}(t_p, \mathcal{L}_2|u_3) = 0.35+0=0.35 \\
&P_{L_0}(t_y|u_3) =  P_{L_0}(t_y, \mathcal{L}_1|u_3) + P_{L_0}(t_y, \mathcal{L}_2|u_3) = 0+0.65=0.65 \\
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_3; t_p) = \log(0.35) = -1.04 \\
&Informativity(u_3; t_y) = \log(0.65) = -0.43
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_1; t_p) = log(P_{L_0}(t_p|u_1)) = \log(1)  = 0\\
&Informativity(u_1; t_y) = log(P_{L_0}(t_y|u_1)) = \log(0) = -\infty
\end{aligned}
$$
We assume that the speaker knows the lexicon, but doesn't believe the listener to know the lexicon. That is, the generative process of the utterance is tantamount to pointing to the table with the referent of the type the speaker wants and incidentally labeling it. The label itself carries no information.

Recall the two lexica: $\mathcal{L}_1 = \{dax: \text{pink-ish thing}, wug: \text{yellow-ish thing}, point: \text{location of point}\}$, $\mathcal{L}_2 = \{dax: \text{yellow-ish thing}, wug: \text{pink-ish thing}, point: \text{location of point}\}$, and the four utterances:  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We begin with $t_p$. If the speaker's lexicon were $\mathcal{L}_1$:

$$
P_{S_1}(u|t_p, \mathcal{L}_1) \propto \exp(\alpha \cdot Informativity(u;t_p)) = \begin{cases} 
\exp(2.24 \times 0) = 1 & \mbox{for } u_1 \\
\exp(2.24 \times -\infty) = 0& \mbox{for } u_2 \\
\exp(2.24 \times -1.04) = 0.097  & \mbox{for } u_3 \\
\exp(2.24 \times -\infty) = 0& \mbox{for } u_4 \\
\end{cases}
$$
To arrive at the production probabilities, you normalize:

$$
P_{S_1}(u|t_p, \mathcal{L}_1) = \begin{cases} 
0.91  & \mbox{for } u_1 \\
0 & \mbox{for } u_2 \\
0.09  & \mbox{for } u_3 \\
0& \mbox{for } u_4 \\
\end{cases}
$$

The speaker's production probabilities are the same if the speaker's lexicon is $\mathcal{L}_2$, only the utterances use the other label:

$$
P_{S_1}(u|t_p, \mathcal{L}_2) = \begin{cases} 
0  & \mbox{for } u_1 \\
0.91 & \mbox{for } u_2 \\
0  & \mbox{for } u_3 \\
0.09& \mbox{for } u_4 \\
\end{cases}
$$
That is, if the speaker is trying to convey the pinkish object type, the speaker would be 10 times more likely (under a speaker rationality parameter of 2.24) to point to the left table than to the right table. 
If instead the speaker wanted to convey the yellowish object type:
 = -0.43

$$
P_{S_1}(u|t_y, \mathcal{L}_1) = \exp(\alpha \cdot Informativity(u;t_y)) = \begin{cases} 
\exp(2.24 \times -\infty) = 0 & \mbox{for } u_1 \\
\exp(2.24 \times -\infty) = 0& \mbox{for } u_2 \\
\exp(2.24 \times -\infty) = 0  & \mbox{for } u_3 \\
\exp(2.24 \times -0.43) = 0.38& \mbox{for } u_4 \\
\end{cases}
$$

To arrive at the production probabilities, you normalize:

$$
P_{S_1}(u|t_y, \mathcal{L}_1) = \begin{cases} 
0  & \mbox{for } u_1 \\
0 & \mbox{for } u_2 \\
0  & \mbox{for } u_3 \\
1& \mbox{for } u_4 \\
\end{cases}
$$
and

$$
P_{S_1}(u|t_y, \mathcal{L}_2) = \begin{cases} 
0  & \mbox{for } u_1 \\
0 & \mbox{for } u_2 \\
1  & \mbox{for } u_3 \\
0& \mbox{for } u_4 \\
\end{cases}
$$

That is, the speaker would point to the right table and say the utterance consistent with their lexicon.

Finally, based on equation 1 we can use these values to compute the probability that the listener thinks that the speaker is referring to the yellow object ($r_1$) when they produce $r_1$:

$$
\begin{aligned}
&P_{L_1}(t, \mathcal{L}|u ) = ...
\end{aligned}
$$
